---
output:
  #bookdown::html_document2: default
  #bookdown::word_document2: default
  bookdown::pdf_document2:
    template: templates/brief_template.tex
    citation_package: biblatex
bib-humanities: true
documentclass: book
bibliography: references.bib
header-includes:
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage[default]{sourcesanspro}
- \usepackage[T1]{fontenc}
- \usepackage{amsmath}
---

# Financial Intermediation and the Transformation of Microfinance Institutions: A synthesis {#cites-and-refs} 
\chaptermark{Transformation of Microfinance: A Synthesis}
\minitoc <!-- this will include a mini table of contents-->

## Introduction
\noindent This chapter provides contextual information that lays the foundation for the review of the empirical literature on the transformation of MFIs. The chapter starts by introducing financial intermediation and then locates microfinance in the financial intermediation space. The section subsequently presents an overview of the evolution of microfinance and the drift towards the conversion of MFIs from the NGO model to the commercial model. Next, the chapter presents the extent of the transformation of MFIs in selected African countries, including the manifestation of the transformation of MFIs. The section concludes by outlining the challenges that MFIs face in the transformation process.

## Need for Financial Intermediation
\noindent Financial intermediaries play a central role in the economy (Kodres, 2017). They contribute to economic development by transforming the maturity and liquidity of assets, providing leverage, and transferring credit risk (Uddin et al., 2014, Valickova et al., 2015, Gajjala and Gajjala, 2016).  By channelling excess funds from the individuals and institutions to those that have deficit savings the intermediaries resolve the problem of information asymmetry between lenders and borrowers and cut costs associated with the duplicated screening of potential borrowers (Greenbaum et al., 2015). 

Typical services provided by financial intermediaries include brokerage and qualitative asset transformation ^[Examples of brokerage services include transaction services (e.g. deposits and cash transfer), financial advice, screening, origination, and funding (loan issuance). On the other hand, qualitative asset transformation involves monitoring, guaranteeing, liquidity creation and the management expertise associated with these activities. See GREENBAUM, S. I., THAKOR, A. V. & BOOT, A. 2015. Contemporary financial intermediation, London, Academic Press. for more details of the services offered by financial intermediaries.] (Greenbaum et al., 2015, Gobat, 2017). Financial intermediation allows individuals to make long-term consumption and investment decisions and overcome liquidity constraints associated with the uncertainty of income streams (Greenbaum et al., 2015). In doing so, financial intermediaries allow for the efficient allocation of capital that improves total factor productivity (International Monetary Fund (IMF), 2017) which contributes to economic growth (Chang et al., 2017). Also, financial intermediaries create money and act as conduits for the transmission of monetary policy (Greenbaum et al., 2015). In turn, financial intermediaries generate revenue primarily from the spread between lending and deposit rates, the sale of securities and fees from customer services (Cosci et al., 2015).

Commercial banks are among the most prominent financial intermediaries. However, there exists a wide array of alternative financial intermediaries which offer services like those of commercial banks, yet they are not commercial banks. The Financial Stability Board (FSB) classified such institutions as shadow banks ^[FSB definition of shadow banking is “the credit intermediation involving entities (fully or partially) outside the regular banking system.”] (Financial Stability Board, 2017) . Examples of shadow banks include credit unions, mutual funds, broker-dealers that finance their assets using repos, insurance companies, finance companies, and MFIs. Central banks do not regulate most shadow banks. For this reason, most shadow banks cannot, for example, get emergency loans from the central bank. Also, deposit insurance does not cover most deposits of shadow banks (Kodres, 2017). 

By their very nature and scale, shadow banks were not a priority for regulators and policymakers until after the 2007 global financial crisis triggered by defaults in the shadow banking system. It was after the crisis that researchers and policymakers realised that the shadow banking sector could be endogenously destabilising, especially with the interconnectedness of the modern global financial system (FSB, 2017). Shadow banks have since been subject to increased regulatory oversight (Adrian and Narain, 2017).
Despite the benefits conferred by the access to and use of financial services, financial intermediation also has its drawbacks. For instance, the profit margin enjoyed by the intermediary has the net effect of increasing the cost of borrowing and reducing the rate of interest on deposits. Consequently, disintermediation, the reliance by large borrowers on direct rather than indirect finance, is gaining popularity (Greenbaum et al., 2015). 

Furthermore, the mainstream financial intermediaries like commercial banks face hurdles in availing financial services to the poor and the financially excluded (including the business ventures these parties pursue ^[Research indicates that small businesses ran by the poor (mostly agricultural-based microenterprises), and SMEs, experience financial exclusion, especially lack of access to credit to finance their operations as well as access to savings products.]). At the base of the pyramid, information opaqueness of the clientele poses a monumental challenge against the provisioning of financial services. For instance, deprived rural dwellers often lack credit histories, a result of being geographically isolated, and financially excluded for prolonged periods (Alimukhamedova et al., 2015). 

Also, individuals in this market segment often lack collateral to secure credit (Brière and Szafarz, 2015). Besides, the tiny amounts of cash transacted at the bottom of the pyramid result in substantial diseconomies of scale (Abate et al., 2014, Cobb et al., 2016). Moreover, the products and services offered by mainstream banks often do not meet the immediate needs of the poor (Gajjala and Gajjala, 2016). Lastly, difficulties in contract enforcement, especially in low-income countries, exacerbate the situation (Klapper and Singer, 2014). Thus, at the bottom of the pyramid, the problem of information asymmetry, lack of tangible assets, and the small scale of operations manifest more acutely.

In the same vein, indigent and financially excluded clients exhibit low levels of literacy and a high incidence of extreme poverty. Consequently, the level of financial illiteracy at the base of the pyramid is similarly high that it hinders the utilisation of financial services, even when they are available (Engström and McKelvie, 2017). Thus, until the advent of microfinance, most of the Microfinance (MF) market segment was not catered for by the mainstream financial intermediaries. Researchers still find that most of the persons in this segment suffer from financial exclusion (Kota, 2007). 

Before the entry of microfinance, informal financing arrangements, like direct financing between family members and friends, family and community merry-go-rounds, self-help groups (SHGs), neighbourhood retail traders and shylocks often filled the void (Klapper and Singer, 2014). These informal mechanisms and alternative formal financial intermediaries like Microfinance Institutions (MFIs) and other shadow banking institutions remain the significant players in offering financial services to the poor and the financially excluded (Klapper and Singer, 2014). In this context, the next section highlights the rise of MF, a significant part of the shadow banking system committed to availing financial services to the poor and the financially excluded.

## Rise of Microfinance
\noindent As already noted, the social and economic empowerment of the poor and the financially excluded, particularly among women and rural dwellers are central to the validation of microfinance (Engström and McKelvie, 2017). The broader view remains that poverty reduction is not only desirable but also has positive externalities. It is in with this hindsight that, in awarding the Nobel Peace Prize to Mohammed Yunus and the Grameen Bank, the Nobel Committee cited the link between poverty reduction and peace ("Announcement of the 2006 Nobel Peace Prize," 2006).

It is paramount to distinguish between MF as a concept, and in reference to the providers of MF services. As a concept, MF refers to the provision of financial services to the financially excluded sections of the population. A large body of institutions and individuals provide the services, including but not limited to MFIs, public and private commercial banks, NGOs, civil society organisations of public interest (OSCIPs). Other providers include international development organisations, commercial retailers, post offices, credit unions, and informal money lenders (Ledgerwood and White, 2006, Marconatto et al., 2016). Nevertheless, the provision of MF has received both support and condemnation from researchers and policymakers.

Despite the reservations about MF, researchers, and policymakers deemed it a useful tool to avail financial services to individuals previously excluded from the financial system (D’Espallier et al., 2017b).  Moreover, given the limitations of the formal banking system, advocates of financial inclusion pushed for microfinance as the viable alternative to efficiently reach the poor and the financially excluded (Chester et al., 2016). Unlike mainstream financial institutions, MFIs reach the poor through innovative approaches such as group lending, regular repayment schedules, progressive lending (dynamic incentives), and collateral substitutes (Bhuiyan et al., 2016). MFIs also focus more on women and people living in rural areas (Ghosh, 2013). 

The model followed by the modern microfinance business is attributable to Mohammad Yunus and Grameen Bank operating in Bangladesh from the mid-1970s (Ghosh, 2013). After the Yunus and Grameen Bank microfinance model gained traction, the number of MFIs operating has exploded across the globe. Table 2.1 gives the current estimated size of the microfinance market based on data from the MIX pooled database. However, not all MFIs avail their data to the database, and hence the data is an under-representation of the scale of the industry. 

Initially, researchers and policymakers hailed MF as the magical apparatus to resolve the problems of development and poverty (Ghosh, 2013). Researchers have associated MF with increased investment by pre-existing business ventures (Newman et al., 2017) and a reduction in gender inequalities (Shahriar and Garg, 2017, Newman et al., 2017, Mafukata et al., 2017, Zhang and Posso, 2017). Further, there is a direct relationship between MF and household welfare (Meador and Fritz, 2017, You, 2013), purchasing power and employment rate (Lopatta and Tchikov, 2016, Raihan et al., 2017). Also, MF supporters contend that it enables families to cope with the effects of climate change (Fenton et al., 2017) and also deal with the shocks arising from natural disasters (Calis et al., 2017). 

Furthermore, researchers have found that access to microcredit leads to higher business start-ups while increasing the profitability of existing businesses (Klapper and Singer, 2014). Other social indicators that have a positive relationship with MF include the savings rate among the poor, health, child mortality, education, and financial inclusion (Shahriar and Garg, 2017, O’Malley and Burke, 2017). At the macro-level, researchers have found an inverse relationship between the size of the MF in a country and poverty levels (Ahlin et al., 2011). Also, research indicates that MF significantly influences economic development by directly increasing the purchasing power of beneficiaries and indirectly via improved capital accumulation and increased employment rate (Lopatta and Tchikov, 2016, Raihan et al., 2017). Given these findings, it was not a surprise that the UN declared 2005 the international year of microcredit followed by the award of the Nobel Peace Prize to Muhammad Yunus and the Grameen Bank in 2006.

\newpage

```{r, echo = FALSE, message=FALSE, warning=FALSE}
library(readxl)
library(tidyverse)
library(kableExtra)

(tribble(~ Region, ~ FSPs, ~ Borrowers_000, ~ `%_Borrowers`, ~ Gross_Loans_000, ~ `%_Gross_Loan`, ~ Depositors, ~ `%_Depositors`, ~ Desposits_USDMillions, ~ `%_Deposits`, "Africa", 193, 5778.2, 0.05, 8489.5, 0.09, 17298.0, 0.18, 9212.10,	0.16, "EAP", 136, 16257.5, 0.14, 15063.7, 0.16, 16117.9, 0.16, 7687.20, 0.13, "ECA", 136, 3082.6, 0.03, 9899.6, 0.11, 5091.0, 0.05, 7664.30, 0.13, "LAC", 345, 22495.3, 0.19, 38843.2, 0.42, 23708.6, 0.24, 27293.10, 0.46, "MENA", 27, 2148.4, 0.02, 1352.9, 0.01, 465.1, 0.00, 251.00, 0.00, "South Asia", 196,	66929.3, 0.57, 18794.1, 0.20, 35109.2, 0.36, 6885.80, 0.12, "Grand Total", 1033, 116691.3, 1.00, 92442.9, 1.00, 98419.8, 1.00, 58993.59, 1.00)) %>% kable(booktabs = TRUE) %>% 
  kable_styling(font_size = 8) %>% landscape()
```

\newpage
However, some researchers view MF as a double-edged sword that is beneficial to individuals and society but also has some unintended consequences (Ganle et al., 2015, Van Rooyen et al., 2012). For example, Ganle et al. (2015) and Van Rooyen et al. (2012) found that while some women indeed get empowered as a result of access to credit, most have little control over the subsequent spending. A significant proportion suffers harassment from MF agents’ for failing to repay the loans. These coercive loan recovery practices ignited the “no pago” protests  movement in Nicaragua ^[In mid-2008, a movement called “movimiento no pago” (Spanish for “a movement for non-payment of loans”) led by over-indebted farmers in Nicaragua protested MFIs operating in the area. The protests were triggered by the arrest of MFI debtors in default. The movement gained political support, including from the president of Nicaragua, Daniel Ortega. In the end, laws were passed that favored the indebted, including the fixing of interest rates at 16% and an extension of loan recovery periods.] and led to several incidences of suicide by MF loanees in Andhra Pradesh , India ^[In 2010, the government of the state of Andhra Pradesh, India, passed an ordinance to protect borrowers from harassment from MFI agents. This came in the backdrop of suicides committed by some indebted MFI clients, protests by MF borrowers, and reports of increasing violence targeted at MFIs. The Andhra Pradesh Microfinance Institutions (Regulation of Money Lending) Ordinance, 2010 requires all MFIs to register with the state government, specify the areas they operate in, the rate of interest and their system of operation and loan recovery. The law also set stiff penalties for coercive loan recovery, prohibited multiple loans to the same borrower, and capped interest chargeable not to exceed the principle loan amount GHOSH, J. 2013. Microfinance and the challenge of financial inclusion for development. Cambridge Journal of Economics, 37, 1203-1219.] (Bastiaensen et al., 2013, Wright et al., 2016). These MF crises resulted in intervention by authorities in both countries, lifting the veil on the dark side of MF. Similarly, although MF does allow women to raise their income levels, in some instances, it has also led to an increase in domestic conflict (Salia et al., 2017).

Furthermore, Newman et al. (2017) uncovered an insignificant link between MF and improvements in consumption, although spending on business-related investment does rise. Shahriar and Garg (2017) also find that the relationship between MF and both business size and profitability only holds in the long run, calling into question the benefits of the short-term loans offered by MFIs. Research has concluded that MF does not often reach the poorest, as they get crowded out by the better-off clients (Abeysekera et al., 2014). Sir Fazle of BRAC, an MF operating in Bangladesh, attributes the failure by the poorest to borrow on stigmatisation by the society, where “fellow villagers viewed them (the poor) as hopeless cases”(Zulfiqar, 2017b).

Likewise, Aguilera et al. (2017) posit that in the short term, some clients of MF benefit more than others. For example, compared to men, women customers of MF get lower profits from their MF funded businesses. On the same line, older women generate fewer profits compared to younger women. Furthermore, Akotey and Adjasi (2016) conclude that the benefits of microcredit are enhanced or sustained only when coupled with microinsurance. Thus, as Onyuma and Shem (2005) note, microcredit alone cannot cure the ills associated with financial exclusion and poverty. The implication is that MF should look beyond microcredit by offering other financial products. Recent developments in MF have seen the introduction of savings products, microinsurance, pension products and cash transfer services (Chandrasekhar and Anantharaman, 2015).

However, there is also a substantial body of research that disputes the benefits of MF (Bateman, 2010a, Bauchet and Morduch, 2013). For instance, the opponents of MF contend that it does not fund genuinely entrepreneurial activity but instead focuses on small retail operations that neither allows the poor to escape poverty nor even have a significant effect on economic growth (Aguilera et al., 2017). MF leads to a glut in the market for similar products by the poor that causes default and overindebtedness. Moreover, the enterprises funded through MF often do not develop linkages with the formal economy, thus inhibiting the development of synergies in productive activities which further limits innovation and productivity improvements (Ghosh, 2013). Furthermore, some researchers posit that MF does not boost employment and education among the rural poor (Bauchet and Morduch, 2013). Others link micro-credit to increased child labour (Hazarika and Sarangi, 2008), increased gender inequalities in access to finance (Zulfiqar, 2017a), and reduced entrepreneurial spirit among the poor (Field et al., 2013). 

Still, some researchers are chiefly concerned with the high-interest rates charged by MFIs and inappropriate lending practices that fail to account for the social, cultural, and economic context of the target clients (Chester et al., 2016). The basis of the lending practices of modern MF is the logic of western capitalism that stresses individual wealth accumulation. By contrast, poor communities emphasise economic processes (not outcomes), shared roles, survival, and varied economic activities which acts as a collective safety net (Chester et al., 2016). 

Researchers have thus argued an over-emphasis on individual achievement, and the advent of group lending is likely to set back social cohesion as members segregate themselves into groups, directly or indirectly monitor each other and impose sanctions on loan defaulters (Chester et al., 2016). Moreover, the logic of western capitalism that forms the basis of MF degrades community ownership and control over assets.  Finally, researchers have argued that the narrow definition of MF may be detrimental to the broader development agenda in poor communities, including the provision of physical infrastructure, security, education and health services (Ghosh, 2013).

Furthermore, some researchers claim that the high-interest rates charged by MFIs are unaffordable by clients, and only serve to trap them in a debt cycle (Wright et al., 2016). Researchers have likened the high-interest rates charged by MF to a poverty penalty, raising ethical concerns (Chang et al., 2017). Moreover, a substantial number of MF clients get credit from several MFIs. The subsequent confiscation of assets further lowers social status (Paprocki, 2016). 

Microcredit was meant to promote the welfare of the poor by funding entrepreneurial activity. However, some researchers have pointed out that it is not reasonable to presume that all poor people desire to or have the capacity to engage in such activity. As Ghosh (2013) asserts, some MF clients pretend they want credit to fund microenterprises but use the cash for subsistence, health care, and school fees. Some even use the loans to buy durable goods like televisions. Scholars assert that it is simplistic to assume that the lack of access to capital is the primary reason behind limiting the engagement of poor people in entrepreneurial activity. As Kimmitt and Munoz (2017) note, capital is a necessary but not sufficient condition for innovation and the uptake of entrepreneurial economic activity, even among the poor. There are additional resources that are complementary to money such as human capital, social capital, and effective institutions that are lacking in most impoverished communities. 

Moreover, even in cases where all these preconditions are present, they may not be enough to trigger structural changes necessary for the reduction of poverty — the environment where these institutions operate matters. Thus, the evaluation of the success or failure of MF must take into account the context of the situation, such as political freedom, economic facilities, social opportunities, transparency guarantees, protective security (Sen, 2014), and even cultural norms (Shahriar and Garg, 2017). It is against this backdrop that some studies recommended a reexamination of the MF business model and called for better regulation of the industry (Johnson, 2013, Ghosh, 2013). Without reforms, conclude Chester et al. (2016), “the MF industry could not only ruin the lives of many borrowers but could also ruin itself.” 

The debate surrounding the contradictory results arising from MF interventions has preoccupied researchers in MF for a prolonged period. Researchers have pointed to the weaknesses in the numerous methods used to evaluate the performance of MFIs (Awaworyi Churchill and Nuhu, 2016), institutional factors (Kimmitt and Munoz, 2017), and diverse cultural norms (Shahriar and Garg, 2017) for the discrepancies.

## Bankrolling MF: The Place of Subsidies and Grants 
\noindent Before the advent of transformation, most MFIs relied on donor grants and subsidies to finance their operations. Grants are private funds from individuals and corporations, whereas subsidies are funds provided by the government to organisations that benefit the public. Although data is hard to come by, some studies estimate that only 23% of MFIs across the globe operate without subsidies and grants (D’Espallier et al., 2013). 

Consultative Group to Assist the Poor (CGAP) (2017) estimates that both public and private donors provided US\$ 2.9 Billion in funding to MFIs in 2013. In comparison, the funds raised by MFI through the sale of debt and equity instruments, and guarantees amounted to US\$ 13.8 Billion, US\$ 3.7 Billion and US$ 1.4 Billion, respectively.  Different funding agencies tend to use various instruments to support MF activities. Bilateral agencies and foundations, for example, mostly use grants as opposed to development finance institutions that typically use debt, only using equity and guarantees sparingly. Multilateral agencies mainly give loans to governments and grants to MFIs for technical assistance. Government mostly offer subsidies to MFIs (Consultative Group to Assist the Poor (CGAP), 2017).

The provision of the grants and subsidies allowed MFIs to reach the poor without the pressure to generate profits to sustain their operations (D’Espallier et al., 2013). Likewise, subsidies are useful for setting up start-ups and risky ventures where alternative funds are not available (Bogan, 2012). The subsidies and grants thus played a significant role in the setting up of the MF industry when the mainstream financial intermediaries did not consider it a viable business venture and commercial funding were unavailable.

As Hudon (2010) note, the presence of subsidies and grants allows the MFIs to charge lower interest rates. However, other researchers posit that MFIs could still manage to charge low-interest rates if they manage their expenses better, even in the absence of subsidies. Further, given that some grants and subsidies support capacity building of MFI staff, the presence of subsidies could indicate better management and operational practices, which have implications on both social and financial performance of the MFIs. On the contrary, some scholars argue that subsidies may encourage lax management leading to inefficiencies (Ghosh, 2013).

Researchers also argue that subsidies and grants serve to repress the rationing role played by interest rates (Hudon, 2010). Interest rates are not only a source of revenue for financial intermediaries but also ensure that only viable projects access financing. It means that a project must generate a rate of return to service the interest payments and make sufficient returns for the investors. The presence of subsidies and grants means that MFIs may fund poor quality projects which reduces social welfare in the aggregate. As Kota (2007) argues, the subsidised MFIs also crowd out the efficient financial intermediaries. In this respect, Ghosh (2013) and Bayai and Ikhide (2016) call for the design of smart subsidies that do not distort the market but rather serve to strengthen the institutions. 

Furthermore, the presence of subsidies in MFIs may push out informal credit suppliers. Although the discussion around informal money lenders is mainly adverse, they still form a significant source of credit for poor households (Klapper and Singer, 2014). The conversation around offering financial services to the poor should then be about how MF can serve to supplement, rather than completely replace the informal financial sector. Instead, the western driven research discourse around MF and poverty dwells on the need to replace the inefficient informal financial systems with an efficient, formal system modelled on western capitalist thinking. The direction taken by researchers fails to reflect on the social-cultural inclinations of the members of poor communities (Chester et al., 2016) and rules out the existence of alternate “systems of exchange” beyond the price-driven model dominant in economics and finance literature (Biggart and Delbridge, 2004).

Lastly, subsidies diminish incentives by MFIs to develop and offer new products. If financing from subsidies is guaranteed, then MFIs lack the motivation to diversify their revenue sources. Indeed, before transformation, most MFIs offered only microcredit (Cozarenco et al., 2016) although legal restrictions did not allow MFIs to provide some products, for instance, savings. After the transformation, the range of the services provided by MFIs expanded. The expansion of the services provided by MFIs post-transformation is, to some extent, the evidence against the reliance on subsidies and grants by MFIs.

## Alternatives and Complements to MF
\noindent Based on the criticism of microfinance, the financial system has witnessed the development and adoption of new strategies to reach the poor. However, these new approaches are not exactly meant to replace MF, but rather to complement it by bridging its shortcomings. A prominent case is the use of mobile money which has proved to be a useful tool for expanding financial inclusion. In Kenya, for example, M-Pesa, M-Kopa, and M-Shwari have contributed positively to financial inclusion by allowing the poor to pay bills, save, transfer and receive money conveniently at a relatively low cost (Klapper and Singer, 2014, Ndung'u et al., 2017). Numerous financial institutions and Fintech companies across the globe have since developed and adopted the mobile money platform. MFIs have espoused mobile money and web technologies to reach their clients (Griffoli, 2017, Ouma et al., 2017). 

“Graduation” program is an alternative scheme where indigent persons are provided with an asset, for example, cattle and offered training in livestock management ^[A graduation program could also involve providing a poor person with seeds, fertilizer, crop insurance and training in plant husbandry. The resources are provided on loan to be repaid at an agreed rate of interest. The One Acre Fund, an NGO based in Eastern Africa, has experimented with this model.] . Results from random controlled trials (RCT) indicate that such “graduation” programs are more successful than the provision of microcredit alone. According to Bishop and Rodríguez (2017), graduation programs result in more wealth and raises the demand for food and other durable goods among the poor people participating in the program. However, the programs are not only expensive but also hard to implement in urban settings where agricultural activity is limited. Furthermore, such wealth generating programs, make products such as savings and insurance even more useful to clients (Cozarenco et al., 2016). 

Finally, governments and other organisations in several developing countries give conditional cash transfers to individuals who meet specific criteria (Rawlings and Rubio, 2005, Tadesse and Zewdie, 2019). For example, a scholarship program operational in Cambodia make cash transfers to households contingent upon the enrollment of children in school (Ferreira et al., 2017). In Kenya, there is a program that awards specified cash awards to citizens above 70 years of age living in poverty (Sandra and Juan, 2017). However, in addition to implementation challenges, conditional cash transfers are affected significantly by political leadership, program administration and even natural disasters (Rodgers, 2014). Moreover, conditional cash transfers have an incentive effect on individuals; where non-poor individuals change their behaviour to receive benefits meant for the poor (Brown et al., 2017). Also, there is mixed evidence regarding the efficacy of cash transfer programs (Sandra and Juan, 2017). 

## MF to Financial Inclusion: The Paradigm Shift
\noindent Recent developments have seen researchers and policymakers start to rethink the central place of MF in achieving financial inclusion (Cobb et al., 2016). The new financial inclusion paradigm recognises that MF is just one element of a broader set of strategies for financial sector development. They call for the development of long-term policies that would enable the players in the financial system beyond MF to avail their services to the poor and the financially excluded (Ghosh, 2013). Thus, although MF has made a valuable contribution to financial inclusion, these researchers argue that a multifaceted approach involving all the alternative providers of financial services could achieve better results (Taylor, 2012). In this respect, almost all financial intermediaries, including commercial banks, have to some extent developed products that suit the needs of the indigent and other financially excluded individuals  (Johnson, 2013, Ndung'u et al., 2017). 

Recommendations to achieve financial inclusion include the development of policies to keep interest rates low and keeping in check to destabilising capital flows (Lund and Harle, 2017). Also useful are legal, fiscal and monetary mechanisms to influence the flow of credit to the poor and the financially excluded. These researchers contend that the achievement of these objectives would require some form of subsidies and grants (Ghosh, 2013).

The financial inclusion paradigm has resulted in heightened competitive pressure among financial intermediaries. Increased competition is forcing MFIs to pay attention to the minimisation of operational costs to boost efficiency. In this respect, MFIs are seeking economies of scale (by netting in more customers) and economies of scope (by rolling out more products beyond microcredit), and leveraging on mobile and web technology and agents in place of brick and mortar branches (Muriu, 2016). Overall, the competitive pressure, the rise of the neoliberal political-economic paradigm, legal requirements and pressure from donors has resulted in some MFIs transforming from NGOs to commercial entities (Muriu, 2016, Kar, 2016). The transformation of MFIs is changing the financial intermediation landscape, even for the indigent clients. The conversion of MFIs is detailed next.

## Transformation of MFIs
\noindent At the initial stages of MF development, most MFIs operated as NGOs, raising capital mainly from the public and private donations. However, starting in 1992, some MFIs have converted to commercial entities sourcing funds in the capital market and emphasising profitability and sustainability, simultaneously with the pursuit of their social mission of serving the poor. Bateman (2010) traces the momentum for the transformation of MFIs from NGOs to commercial entities to the rise of neo-liberalism in the mid-1970s. Core to the neo-liberal paradigm was the need for all institutions operating in the economy to be financially self-sustainable. Thus, economic liberalisation, commercialisation, and privatisation of public enterprises characterised the neo-liberalism wave. 

In developing countries where most of the formal MFIs operate, the IMF and the World Bank induced structural adjustment programs (SAPs) were the significant indicators of the global reach of neo-liberalism. SAPs were designed to allow the economies of target countries to be more market-oriented (Easterly, 2003). Similarly, the World Bank, USAID, and other donor agencies pushed for the transformation of MFIs with the aim of expanding MFI outreach without the need for subsidisation (Ghosh, 2013). For MFIs, the adjustments meant that the interest rates charged were to be determined by market forces in place of the below-market, subsidy-supported rates (Bateman, 2010a). Similarly, MFIs that undertook the transformation worked to increase the client base to lower the fixed cost per client and offered incentives to senior managers, including significant ownership stakes in the MFIs (Lauer, 2008). In some other countries, for example, Georgia , legal changes forced MFIs to transform into commercial entities ^[Legislation enacted in Georgia required that all NGOs cease engaging in MF activities by December 31, 2007 LAUER, K. 2008. Transforming NGO MFIs: Critical ownership issues to consider, New York, Consultative group to assist the poor (CGAP)].

With neo-liberalism, most of the donor organisations that channelled funding to MFIs also pushed for the commercialisation of the MFIs. To date, a substantial number of MFIs have converted to commercial entities. Notable though is that most MFIs have survived the neo-liberal onslaught and still operate as NGOs (D’Espallier, Goedecke, Hudon, & Mersland, 2017). The shift from the NGO model to the commercial model has several implications. First, researchers point out that the profit motive has the potential to shift the focus of MFI’s away from their social mission of serving the poor and the financially excluded to the emphasis on financial returns. “Mission drift” is the term used to describe this potential change in business model focus (Mersland & Strøm, 2010). 

Similarly, the departure from predominant reliance on donor funding to commercial funding has implications from the agency theory perspective. In this respect, agency conflicts may manifest between shareholders on the one hand, and the management, debt-holders, private and public donors on the other, and this influences the performance of the MFI (Fama, 1980). Also, agency conflicts have the potential to affect the funding structure and performance of MFIs (Berger and Bonaccorsi di Patti, 2006, Le and Phan, 2017). Consequently, researchers have raised questions regarding how well MFIs have balanced financial self-sufficiency and the pursuit of social goals after the shift to the new business model. It is upon this basis that the world bank launched an equity fund to support MFIs to meet the costs of the transformation and to safeguard their social mission (Frank et al., 2008).

## Schools of Thought on the Transformation of MFIs
\noindent As noted, state agencies, donor agencies, and philanthropists were the most significant sources of funding for MFIs (Ledgerwood, 1998, Ledgerwood and White, 2006). The initial thrust for MF arose from the desire to lift the billions of people that languished in poverty, mostly in third world countries. Availing credit to the poor was and is still seen as an essential first step towards breaking the vicious cycle of poverty. However, some researchers contest this view (Bateman, 2010a, Adams and Vogel, 2016, Onyuma and Shem, 2005). Thus, MFIs were established solely to tackle human misery that arose from the scourge of poverty. 

In the circumstances that motivated the establishment of the first MFIs like Grameen, the profit motive was ethically incompatible with the MF agenda (Ghosh, 2013, La Torre, 2006) due to the ethical concern around the “financialization” of poverty (Mader, 2016). Yunus had emphasised the need for MFIs to maintain low interest rates, in contrast with the usurious rates charged by moneylenders in rural Bangladesh that Grameen Bank and other MFIs were supposed to replace (Bateman, 2010a). Consequently, MFIs set the interest rate below the market rate. Accordingly, MFIs operated as NGOs, focusing more on their social mission of serving the poor and the financially excluded than the generation of profits. The welfare approach to the provision of MF underpinned this model.

Proponents of the welfare approach argued that MFIs could attain their social mission without being financially self-sufficient (Kodongo and Kendi, 2013a). The welfare approach further holds that the social purpose of MFIs is paramount and incompatible with financial sustainability. To the researchers who subscribe to this school, financial viability (commercial orientation) and an emphasis on the social mission (development positioning) of MFIs are different approaches that have different aims and motives (Toindepi, 2016). They cite ample body of empirical studies that uncover a decline in social performance by MFIs that have to pursue both the profit motive and the social mission (i.e., transformed) in support of their view. Bos and Millone (2015) and Mia and Lee (2017), for instance, find that capital market funding is subject to mission drift, the evidence against the feasibility of the institutional transformation of MFIs.

Over time, however, some researchers and policymakers began to question the feasibility of the welfare approach. Their primary concern was that the reliance by MFIs on donor funding, though socially desirable, was not a viable strategy.  The result was the institutional sustainability school, an antithesis of the welfare school. The new school championed the financial self-sufficiency of MFIs, a stand that raised ethical concerns (Mersland and Strøm, 2010). Despite the attacks from the moral standpoint, researchers pointed out that sustainability confers advantages to MFIs that outweighs the potential drawbacks. 
First, sustainability allows MFIs to be able to source capital from the financial markets and hence improve control and governance, and customer service (see section 1.3). Likewise, the diversity in funding sources permits the extension of financial services to a broader group that would not be possible under the limited financing available to NGOs. Moreover, the diversity of capital sources cushioned the MFIs from global economic shocks and political uncertainties that affect the flow of donor funds (Garmaise and Natividad, 2013, D’Espallier et al., 2017b). Indeed, some research is supportive of the sustainability school. For example, Tchakoute-Tchuigoua (2010) found that private driven MFIs perform better financially when performance is measured using portfolio quality. They further documented that for-profit MFIs are more socially efficient as compared to their NGO counterparts. Besides, there is also evidence that transformed MFIs score better regarding general outreach, growth in loan portfolio, outreach to women, and the range of products evolved (Frank et al., 2008). 

However, there are other studies which have found that the two paradigms can coexist without significant loss of social functionality (Mersland and Strøm, 2009, Mersland and Strøm, 2010).  This compromise view, called the win-win approach, is an attempt to reconcile the social mission school and the institutional sustainability approach. This school asserts that it is possible to balance sustainability with the attainment of social goals (Kodongo and Kendi, 2013a). A prominent study in support of this paradigm is Louis et al. (2013) who uncover a positive relationship between social efficiency and financial performance, using data of 650 MFIs across the globe. Similar results from a study based on 31 MFIs in Africa,  and applying the Generalized Structural Equation Models, found a direct positive effect (and an indirect adverse effect) running from viability to depth of outreach (Ayele, 2015).

Despite the different schools, the MFI transformation movement appears to have taken a firm footing in the MF industry. Thus, the discussion around the merits and demerits of transformation, though necessary, may not be addressing the new reality. To resolve the challenges posed by the commercialisation of MFIs, the output from studies that examine the effects of the novel MFI organisational structures on the social versus financial performance comparison of MFIs would be useful. Such studies would also offer solutions and policy direction that will guide the future path of this vital sector. The next section highlights the current state of the MF industry in Africa, including the extent of the conversion of MFIs in Africa from NGOs to commercial entities. 

## Microfinance in Africa and the Extent of the Transformation of MFIs 
\noindent The financial sector in Africa is still relatively shallow by global standards and trails the world in size, financial sustainability, and outreach (Azad et al., 2016). African countries saw a five-fold increase in financial access, a 40% increase in GDP per capita and an explosion of mobile money (Klapper and Singer, 2014, Beck and Cull, 2014). Still, less than one in five individuals in the continent have access to financial services, although there are significant regional disparities  (International Monetary Fund (IMF), 2017, Beck and Cull, 2014). Microfinance and the informal financial sector have played a crucial role in availing financial services to the poor and the financially excluded (Klapper and Singer, 2014, Engström and McKelvie, 2017). 

In Africa, several MFIs have converted to commercial entities, the first being K-REP in Kenya in 1999. However, most MFIs in Africa still operate as NGOs, or a hybrid of NGO and commercial model, meaning that donations form a substantial part of the capital of MFIs (Bayai and Ikhide, 2016). The next section provides an outline of the MF industry in Africa and also a synopsis of the extent of the transformation of MFIs in selected African countries. 

### Microfinance in Africa: An Overview
\noindent Given its geographical size, cultural diversity, varied colonial heritage, and differing levels of social-economic development, microfinance in Africa is hardly homogeneous. The cross-country and regional differences manifest both in the nature and levels of development of the MF industry across the continent. For instance, the MF industry in North Africa is closer to that of countries in the Middle East given the shared language and religion. Thus, although the countries in North Africa are members of the African Union, the Microfinance Information Exchange (MIX) categorises North Africa together with the Middle East (as the MENA region). The discussion on microfinance that follows mainly dwells on the Sub-Sahara African (SSA) countries. Where available, though, data from North Africa are also incorporated. Overall, the data are just an estimate given the large informal financial market in Africa (Klapper and Singer, 2014). 

Undoubtedly, the most extensive MF market in Africa is Nigeria both regarding borrowers and depositors. By 2015, Nigeria had 1,698,800 borrowers and a gross loan portfolio of US\$ 542,500,000. However, Kenya had the highest gross loan portfolio of US$3,290,400,000 against 374,000 borrowers indicating a higher average loan size. Figure 2.1 (Panel A) below is a visualisation of the relative size of MFIs in Africa. 

True to its purpose, Microfinance in Africa lends primarily to Microenterprises, SMEs, and households. Figure 2.2, panels C shows that the bulk of the loans by count goes to microenterprises (84.6%) and household financing (11.5%). SMEs hold 3.5% of the loans, whereas large corporations have a negligible 0.4% of the loans. Note that these statistics are counts of individual loans granted, not monetary worth. What is in dispute is whether MFIs in Africa and indeed the world over, reach the poorest of the poor. Thus, although most of the lending goes to microenterprises, it is not clear whether the poorest and the financially excluded own these microenterprises (Beck and Cull, 2014). Moreover, even where the welfare of the poor and the financially excluded has improved, it is not clear whether such improvements are a result of MF interventions (Awaworyi Churchill and Nuhu, 2016).



```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(patchwork)
#######################Gross Loans/ Borrowers in Africa
Countries <- c("Nigeria", "Benin", "Kenya", "Ghana", "Uganda", "Morocco", "Egypt", "Tunisia")

Borrowers <- c(1698.8, 462.3, 374, 281.4, 223.3, 763.6, 639.5, 270.6)

Gross_loans_USD_Millions <- c(542.5, 127, 3290.4, 507.8, 363.3, 566.4, 157.6, 146.4)

borrowers_loans <- data.frame(Countries, Borrowers, Gross_loans_USD_Millions)

############## Deposits and depositors in Africa
Countries <- c("Nigeria", "Burkina Faso", "Kenya", "Ghana", "Uganda", "Morocco", "Egypt", "Tunisia")

Depositors <- c(3396.4, 1172.0, 1571.0, 1774.5, 1648.0, 0, 0, 0)

Gross_Deposits_USD_Millions <- c(260.3, 222.9, 3632.0, 727.4, 472.1, 0, 0, 0) 

depositors_deposits <- data.frame(Countries, Depositors, Gross_Deposits_USD_Millions)

################# Number of loans by sector
Sector <- c("Household_financing", "Large_corporations", "SMEs", "Microenterprise")

Proportion <- c(11.5, 0.4, 3.5, 84.6)

Loans_by_use <- data.frame(Sector, Proportion) %>% 
  
  mutate(Sector = factor(Sector))


################### Gross loans by sector
Sector_gross <- c("Household_financing", "Large_corporations", "SMEs", "Microenterprise")

Proportion_gross <- c(37.6, 14, 12.7, 35.7)

Loans_by_use <- data.frame(Sector_gross, Proportion_gross) %>% 
  
  mutate(Sector = factor(Sector_gross))

gross_loans_by_use <- data.frame(Sector_gross, Proportion_gross)

###################### Average loan balance by sector
Average_loan <- c(2974, 29238, 3329, 382)

average_loan_data <- data.frame(Sector, Average_loan)

##################### Number of offices by MFI legal status in Africa
Legal_status <- c("Bank", "Cooperative", "NBFI", "NGO", "Rural Bank")

Number_of_offices <- c(816, 1854, 915, 1313, 9)

Borrowers_per_loan_officer <- c(343, 715, 285, 580, 1294)

Legal_status_data <- data.frame(Legal_status, Number_of_offices, Borrowers_per_loan_officer)

####################### Number of MFIs by legal type
No_of_MFIs <- c(30, 67, 57, 36, 2)

data_no_of_mfis <- data.frame(Legal_status, No_of_MFIs)

###################### Rural urban service delivery
rural_urban <- data.frame(Area = c("Rural", "Urban"), Prop = c(64, 36))

```




\newpage
\blandscape
```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 9.5, fig.height=5.9, fig.cap = "Metrics on MFIs Africa"}

#############################Graphs
######################## No of MFIs
(data_no_of_mfis %>% 
  
  mutate(Legal_status = factor(Legal_status)) %>% 
  
  ggplot(aes(x = fct_reorder(Legal_status, No_of_MFIs, max), y = No_of_MFIs, group = Legal_status)) + 
  
  geom_col(show.legend = FALSE) +
  
  geom_label(aes(label = No_of_MFIs), show.legend = FALSE) +
  
  ggthemes::theme_clean() + 
   
   scale_fill_brewer(palette = "Dark2") +
  
  labs(x = "Legal Status", y = "Number", title = "A: Number of MFIs by Legal Status in Africa- MIX Dataset") + 
   
   theme(plot.title = element_text(size = 8)) +
  
  coord_flip() +

############### Deposits and depositors
depositors_deposits %>% 
  
  pivot_longer(-Countries, names_to = "Metric", values_to = "Amount") %>% 
  
  ggplot(mapping = aes(x = fct_reorder(Countries, Amount, max), y = Amount, fill = Metric)) + 
  
  geom_col(position = "dodge") + 
  
  ggthemes::theme_clean() + 
  
  theme(legend.position = "bottom", legend.title = element_blank(), legend.text = element_text(size = 7), plot.title = element_text(size = 8)) + 
  
  labs(x = "Country", y = "Deposits/ Depositors", title = "B: Deposits and Depositors in Africa") + 
  
  coord_flip() + 
  
  scale_fill_brewer(palette = "Dark2")) /

######################Rural urban
(rural_urban %>% 
  
  ggplot(aes(x = fct_reorder(Area, Prop, max), y = Prop, group = Area)) + 
  
  geom_col() + 
  
  geom_label(aes(label = Prop)) +

  ggthemes::theme_clean() + 
  
  labs(x = "Area (Rural/Urban)", y = "Proportion(%)", 
       
       title = "C: Rural vs Urban Clients Served by MFIs in Africa") + 
  
  coord_flip() + 
  
  scale_fill_brewer(palette = "Dark2") +
  
  theme(legend.position = "bottom", legend.title = element_blank(), legend.text = element_text(size = 7), plot.title = element_text(size = 8)) +

################################ Borrowers by Legal status data
Legal_status_data %>% 
  
  pivot_longer(-Legal_status, names_to = "variable", values_to = "values") %>% 
  
  mutate(Legal_status = factor(Legal_status)) %>% 
  
  ggplot(mapping = aes(x = fct_reorder(Legal_status, values, max), y = values, fill = variable)) + 
  
  geom_col(position = "dodge") + 
  
  geom_label(aes(label = values)) +
  
  ggthemes::theme_clean() + 
  
  labs(x = "Legal Status", y = "", 
       
       title = "D: Borrowers per Loan Officer and Number of Offices") + 
  
  coord_flip() + 
  
  scale_fill_brewer(palette = "Dark2") +
  
  theme(legend.position = "bottom", legend.title = element_blank(), legend.text = element_text(size = 7), 
        
        plot.title = element_text(size = 8))) + 
  
  plot_annotation(title = "Metrics on MFIs Africa", caption = "Source: Construction from MIX data")

```




\newpage
```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 9.5, fig.height=5.9, fig.cap = "Metrics on MFI Loans in Africa"}

########################Gross loans and borrowers
(borrowers_loans %>% 
  
  pivot_longer(-Countries, names_to = "Metric", values_to = "Amount") %>% 
  
  ggplot(mapping = aes(x = fct_reorder(Countries, Amount, max), y = Amount, fill = Metric)) + 
  
  geom_col(position = "dodge") + 
  
  ggthemes::theme_clean() + 
  
  theme(legend.position = "bottom", legend.title = element_blank(), legend.text = element_text(size = 7), 
        
        plot.title = element_text(size = 8)) + 
  
  labs(x = "Country", y = "Borrowers/ Gross Loans", title = "A: Gross Loans and Borrowers in Africa") + 
  
  coord_flip() + 
  
  scale_fill_brewer(palette = "Dark2") +

############################ Gross loans by sector
gross_loans_by_use %>% 
  
  mutate(Sector = fct_recode(Sector_gross, "Household" = "Household_financing")) %>% 
  
  mutate(Sector = fct_recode(Sector_gross, "Large_corp" = "Large_corporations")) %>%
  
  ggplot(aes(x = fct_reorder(Sector_gross, Proportion_gross, max), y = Proportion_gross)) + 
  
  geom_col() + geom_label(aes(label = Proportion_gross), label.size = 0.15) + ggthemes::theme_clean() + 
  
  labs(x = "Sector(%)", y = "Proportion", title = "B: Gross MFI Loans by Sectors in Africa (%)") + 
  
  theme(plot.title = element_text(size = 8)) +
  
  coord_flip() + 
  
  scale_fill_brewer(palette = 2)) /

####################### Number of loans by sector
(Loans_by_use %>% 
   
  mutate(Sector = fct_recode(Sector_gross, "Household" = "Household_financing")) %>% 
  
  mutate(Sector = fct_recode(Sector_gross, "Large_corp" = "Large_corporations")) %>%
  
  ggplot(mapping = aes(x = fct_reorder(Sector, Proportion, max), y = Proportion)) + geom_col() + 
  
  geom_label(aes(label = Proportion), label.size = 0.15) + 
  
  ggthemes::theme_clean() + 
   
   theme(plot.title = element_text(size = 8)) +
  
  labs(x = "Sector(%)", y = "Proportion", title = "C: Number of MFI Loans by Sectors in Africa (%)") + 
  
  coord_flip() + 
  
  scale_fill_brewer(palette = 2) + 

######################## Average loan data
average_loan_data %>% 
  
  mutate(Sector = factor(Sector)) %>% 
  
  mutate(Sector = fct_recode(Sector, "Household" = "Household_financing")) %>% 
  
  mutate(Sector = fct_recode(Sector, "Large_corp" = "Large_corporations")) %>%
  
  ggplot(aes(x = fct_reorder(Sector, Average_loan, max), y = Average_loan, group = Sector)) + 
  
  geom_col(show.legend = FALSE, col = "black") +
  
  geom_label(aes(label = Average_loan), show.legend = FALSE, label.size = 0.15) +
  
  ggthemes::theme_clean() + 
  
  labs(x = "Sector(%)", y = "Average Loan Balance", 
       
       title = "D: Average MFI Loans by Sector in Africa (US$)") + 
  
  coord_flip() + 
  
  scale_fill_brewer(palette = "Dark2") +
  
  theme(plot.title = element_text(size = 8))) +
  
  plot_annotation(title = "Metrics on MFI Loans in Africa", caption = "Source: Construction from MIX data")

```



\elandscape
\newpage
To illustrate the difficulty that researchers face in evaluating the impact of MFIs on the poorest segments of the society, consider the following two cases. Pitt and Khandker (1998) applied the Randomized Control Trials (RCTs) and found that group-based lending benefits the poor. However, when Roodman and Morduch (2014) replicated the study by applying a robust linear estimator and dropping outliers, they found that the results do not hold. Ghosh (2013) also replicated an evaluation report by USAID of the Self-Employed Women’s Empowerment (SEWA) based in India and found most of the positive evaluations not to hold after factoring in individual characteristics of the participants. It is thus difficult to arrive at a firm conclusion that MF positively impacts on the poor given the methodological weaknesses inherent in MF impact evaluation methods (Awaworyi Churchill and Nuhu, 2016). 

For the total gross loan portfolio, a similar picture emerges, although the distribution is not as extreme as that of the loan counts above. Figure 2.2 panel B illustrates that household financing and microenterprises hold the highest shares of the gross loan portfolios at 37.6% and 35.7%. The share of the gross loan portfolio held by large corporations and SMEs rises to 14% and 12.7%, respectively. Thus, although microenterprises and households produce the highest number of borrowers, the average loan amount lent to microenterprises and households is much smaller compared to SMEs and large corporations. The distribution may be an indicator or confirmation that indeed MFIs may not reach the poorest given that large enterprises receive only 0.4 % of the loans by count but which amounts to 14% of the gross loan portfolio. 

Figure 2.2 panel D shows the average loan balances held by microenterprises, households, SMEs and large enterprises. Although large enterprises hold 0.4% of the loans granted by count, the average loan size is 77 times larger than that of a microenterprise. 
Similarly, SMEs, with 3.5% of loans by count, have an average loan size that is nine times larger than that of a microenterprise. Thus, many microenterprises and households borrow relatively smaller amounts compared to SMEs and large enterprises. Again, it is not clear whether the loans do reach individuals at the very bottom of the pyramid where households subsist on less than a dollar a day.

### Transformation of MFIs in Selected Africa Countries
\noindent This section highlights the extent of the transformation of MFIs in Africa. The section draws on the available data from the MIX pooled database and the IMF 2017 Financial Access Survey database. The study draws a sample of 2 countries from each of Africa’s five regions based on the number of MFIs submitting data to the MIX pooled database (at least 5). Next, the relative sizes of the gross loan portfolio and deposits form the basis of the selection. However, the study also includes Ethiopia ^[In Eastern Africa, three (3) countries are examined. The addition of Ethiopia is based on the fact that it is among the African countries with the lowest level of capital market development, and notably lacking a stock market THE OFFICIAL MONETARY AND FINANCIAL INSTITUTIONS FORUM 2017. Barclays Africa Group Financial Markets Index 2017. Johannesburg: Barclays Africa Group Limited.. However, Ethiopia also has an extensive MF sector with substantial state involvement AYELE, G. T. 2015. Microfinance Institutions in Ethiopia, Kenya and Uganda: Loan Outreach to the Poor and the Quest for Financial Viability. African Development Review, 27, 117-129.. It is the view of the researcher that Ethiopia would be a useful addition to the study] in the Eastern African region. Table 2.3 shows the summary statistics relating to MFIs in the selected countries. Note that for Southern Africa, only South Africa meets the criteria. Botswana, Namibia, Lesotho, and Swaziland have no MFIs submitting data to the MIX pooled database.

\newpage
\blandscape

```{r, echo = FALSE, message = FALSE, warning = FALSE, results='asis'}
library(tidyverse)
library(kableExtra)
regional_mfi_data <- read.csv("regional_statistics.csv") %>% 
  
  rename(Gross_Loans_USD_Millions = Gross.Loan.Portfolio.US..Millions..2015., Average_Loan_USD = Average.Loan.per.Borrower.US., Desposits_USD_Millions = Deposits.US..Millions..2015.)

regional_mfi_data %>% 
  
  kableExtra::kbl(., caption = "Regional Statistics for MFIs, Gross Loan Portfolio, and Deposits (2015)", 
      
  booktabs = TRUE) %>% 
  
  kable_paper(full_width = FALSE, font_size = 8) %>% 
  
  footnote(general = "Microfinance Information Exchange, MIX (2017)",
           
  general_title = "Source: ", 
  
  number = c("The Global Outreach & Financial Performance Benchmark Report – 2015 is the chief source of the data", "However, the MIX database on http://www.themix.org/mixmarket/countries-regions is the source of the number of FSPs.", "The study uses the African Union classification system on regions of Africa"),
  
  number_title = "Note: "
  )

```

\elandscape


## Manifestation of the Transformation of MFIs
\noindent The transformation of MFIs from NGOs to commercial entities has manifested itself in several ways. These include increased reliance on commercial capital, the advent of regulation, and a broader range of financial services on offer. The next sections describe these manifestations.

### Reliance on Commercial Sources of Capital
\noindent As already noted, MFIs relied on grants and subsidies to finance their operations. However, starting in the 1990s when MFIs began to convert to formal organisations they began sourcing capital from commercial sources. However, the presence of commercial capital means that MFIs had to also strive for financial returns, given that research indicates that commercial logic drives the private funders. 

Commercial capital is accessible to transformed MFIs since they appear more creditworthy. The perception of creditworthiness is especially crucial given that most investors favour the governance, risk management and regulatory oversight associated with transformed MFIs when assessing the risk-return profile of investments (Frank et al., 2008). Consequently, most of the transformed MFIs issued financial assets (debt and equity) and mobilised deposits to raise additional capital. Some MFIs are even securitising their loan portfolios, for example, ICICI (India) in 2003, the first MFI to do so (Dieckmann et al., 2007). A few MFIs have even floated their shares through IPOs, such as SKS in India in 2010 (Chen et al., 2010) and the Mexican MFI Banco Compartamos in 2007 (Ashta and Hudon, 2012). However, some researchers have disparaged the IPOs as a mechanism of self-enrichment by a few at the expense of the poor (Ghosh, 2013).

The transformation was a vital step in accessing commercial capital given that, for example, most jurisdictions do not allow non-regulated entities to collect deposits or sell financial assets (Dieckmann et al., 2007). Availability of commercial capital allowed MFIs to free themselves from capital constraints by allowing MFIs to diversify the sources of capital beyond donor funds. Also, research shows that MFIs which mobilise savings are less vulnerable to adverse market conditions (Frank et al., 2008). Likewise, donor funds are negatively prone to economic downturns and politics (Garmaise and Natividad, 2013). However, the presence of commercial capital implies that the MFIs must generate enough revenue to not only meet their social goals but to guarantee the expected financial return to the providers of capital. The presence of a financial objective and the social mission means that MFIs must optimise a double bottom line (Mersland and Strøm, 2010). MFIs must also manage their cost of capital, and deal with the agency costs arising from the entry of shareholders and bondholders.

Nevertheless, the shift to commercial model is no guarantee that funding from the capital markets will necessarily flow to transformed MFIs, Factors such as a country’s legal tradition, creditor rights, and financial sector development are important determinants of the flow of formal external financing to MFIs (Tchakoute-Tchuigoua, 2014). On the other hand, the entry of commercial funding could shift the focus of MFIs away from their social mandate (Casselman and Sama, 2013). Much of the debate on transformation revolves around the effects that this shift has had on the social and financial performance of MFIs. 

Bogan (2012) notes that the source of capital could influence sustainability and social mission attainment by MFIs. The roots of the argument are that funding source is directly linked to the level of monitoring and hence may incentivise or discourage efficiency. Other factors such as the regulatory environment influence the capacity of an MFI to attract certain classes of capital. Table 2.4 below highlights the relative merits and demerits of various sources of funding considering the agency theory.
It is vital to note that most of the transformed MFIs operate on a mix of various capital sources. The most common combination is that of debt, deposits, equity, and grants. Given the merits and demerits of each source, it is not clear how the influences associated with each source interact and shape the performance of MFIs, both socially and financially. Most researchers have adopted a dichotomy-based approach to the study of the sources. For example, some studies compare firms with mostly commercial funding vis-à-vis those with grants. In some other studies, researchers based their analysis on the source of funding, public or private.

\blandscape

```{r, echo = FALSE, message = FALSE, warning = FALSE, results='asis'}
library(readr)
regional_statistics2 <- read_csv("regional_statistics2.csv")

regional_statistics2 %>% 
  
  kableExtra::kbl(., caption = "Funding Sources and Critique)", 
      
  booktabs = TRUE) %>% 
  
  kable_paper(full_width = TRUE, font_size = 8) %>% 
  
  footnote(general = "Adapted from Bogan (2012)",
           
  general_title = "Source: "
  ) 
```

\elandscape

Beyond, monitoring, the regulatory environment may determine the sources of funding. Capital sources such as deposits, public debt, and equity are subject to regulatory approval and oversight. The next section examines the rise of regulation for MFIs.

### Regulation
\noindent The financial markets legislation “defines the rules, controls, and sanctions which restrict and enable the behaviour of participants in financial markets” (Marti and Scherer, 2016). Legislation targeted at the MFI sector has risen concomitant with the transition from the NGO model to the commercial model. The regulation is meant to maintain stability in the MF sector (and the economy in case of spillovers), encourage competition, protect depositors and borrowers, and maintain the payment systems (Johnson, 2013). The laws targeted at the MF sector were in response to several negative experiences among MFIs across the globe, and also due to the pressure from MFIs and other interest groups for the development of MFI specific regulations (Campion and White, 1999). 
In Bangladesh, for instance, indigent clients lost their savings due to the incompetence and fraud by some MFIs (Paprocki, 2016). In Nicaragua, the “no pago” movements raised concerns that MFIs could harm communities by trapping them in unsustainable debt (Bastiaensen et al., 2013). Moreover, cases of suicide by over-indebted MFI clients in India added to the ethical concern around MF practices (Roodman, 2012). Furthermore, a service such as deposit-taking and the entry of commercial funds meant that regulation to protect depositors and other stakeholders from the ownership and governance, management, portfolio, and new industry risks associated with commercially oriented MFIs (Meagher et al., 2006) was necessary. Finally, the 2007 global financial crisis served to illustrate that even the shadow banking system can generate economic shocks that can affect the entire financial system and indeed the economy (Meeks et al., 2017).

In the absence of MFI specific legislation, the initial MFIs that transformed did so under commercial banking laws (Campion and White, 1999, Siwale and Okoye, 2017). However, numerous countries have developed legislation specific to MF. Examples of countries that have MFI specific legislation include Kenya (2006), Bolivia (starting from 1958 to 2002)^[There are several laws that govern MFIs in Bolivia. These include the Law on Banks and Financial Entities (1993), Central Bank of Bolivia Law (1995), Property & Popular Credit Law (1998), Law for Normative Strengthening & Financial Supervision (2001), and the Solidary Bond Law (2002). These laws apply to all private financial funds (FFPs), open savings and loan cooperatives (CACs), and commercial banks. CACs are also governed by the General Law for Cooperative Associations (1958) together with Supreme Decree No. 24439 (1996). FFPs are also subject to the Supreme Decree No. 24000 (1995).], Nepal (1996, 1999, 2002), and Indonesia (1992, 2003) (Meagher et al., 2006). In most of these countries, the provisions in MF laws mostly follow a variant of the CAMELS rating model used in bank supervision or a variant of the model. The rules emphasise minimum capital and capital adequacy requirements, asset quality, management quality, earnings liquidity, and the sensitivity to market risk (Daher and Le Saout, 2013). 

However, regulation is costly both socially and financially both to the MFIs and to the regulators. The costs arise mainly due to the involvement of a significant proportion of skilled labour necessary for effective regulatory compliance (Cull et al., 2011). Also, as Marti and Scherer (2016) argue most jurisdictions follow the technocratic approach to the development of financial regulation. Unlike the inclusive approach to legislation, the technocratic approach ignores the affected parties (like the MFI clients), focusing instead on experts and special interest groups. The inclusive approach favours both the means and the ends of financial regulation. Given the contested nature of poverty and financial exclusion, especially the approaches to tackle them, the researchers recommend the inclusive approach.

Finally, research has not found compelling results regarding the benefits of regulation on both the financial and social performance of MFIs. For example, Cull et al. (2011) demonstrate that regulatory supervision leads to reduced outreach to women and other customers that are costly to reach. Similarly, Siwale and Okoye (2017) do not find evidence that regulation of MFIs helps augment the social mission of MFIs, just as Hartarska and Nadolnyak (2007) find no evidence that regulation positively influences both financial and social performance. The researchers do note, however, that regulation has restored confidence in the MFI sector and served to professionalise the provision of MF services.

### Broader Range of Services Offered by MFIs
\noindent Commercialisation saw the reduction in donor funding and the push by commercial capital providers for a financial return, in addition to the pursuit of their social mission. Further, the commercialisation also led to increased competition between commercial banks and MFIs at the base of the pyramid. MFIs adapted to the increased competition and demands from the providers of capital by offering additional products tailored to meet the needs of the poor and the financially excluded (Cozarenco et al., 2016). The rollout of the new products was also meant to increase customer loyalty, attract new customers, reduce institutional risks and improve competitiveness (Frankiewicz and Churchill, 2011). Thus, the product range has expanded to include savings products, microinsurance, cash transfer services, pension accounts, leasing, grants, and a range of non-financial services^[The non-financial services offered by MFIs include health education, skills training, and social awareness.].

The microsavings product is similar to the savings products offered by mainstream commercial banks, credit unions, and building societies. The only significant distinction is the scale of the savings, with MFIs dealing mainly with small savings garnered from the needy individuals. Although the initial assumption was that the poor could not save due to their limited income, research indicates that the indigent clients demand both credit and savings products simultaneously (Afzal et al., 2015). However, the study concludes that the poor are a different lot and thus it is not enough to avail a savings product to them. MFIs must avail the right savings product along with facilitation as the standard savings product offered by commercial banks may not get the poor to save (Chandrasekhar and Anantharaman, 2015).

Money transfer, on the other hand, allows one party in a transaction to move cash to another individual or institution in a scenario where the physical exchange of cash is not possible. For example, M-Pesa is a popular mobile phone-based platform in Kenya that people use to pay utility bills, school fees, and even for government services (Maake et al., 2014). Before the advent of MF, such services usually involved the use of a money order drawn in a post office at substantial cost and inconvenience to the parties involved. The other mechanisms for cash transfer offered by MFIs include checks and bank drafts, electronic fund transfers (EFT), and services based on proprietary networks such as the Western Union and Moneygram (Ouma et al., 2017). Given that MFIs are accessible to the poor, they are in a better position to offer these services at the base of the pyramid.

Microinsurance, or insurance for the poor, is aimed at cushioning the poor against unexpected shocks. The low uptake of the conventional insurance products has led to the push for this model of risk mitigation. Unlike microcredit, microinsurance involves the poor making regular premium payments and cannot quickly be packaged as a group product as has been the case with the microcredit group schemes (Platteau et al., 2017). Also, Armendáriz and Morduch (2010) argue that insurance for the poor is risky given the prevalence in poor communities of infectious diseases and higher susceptibility to natural disasters like droughts and floods. Also, information problems, higher per unit transaction costs of insurance, and difficulties in contract enforcement in poor communities make microinsurance especially perilous and needing to be creatively designed.

Thus, in designing microinsurance products, MFIs must take cognisance of the precarious conditions in which the poor live. To encourage the poor to cover themselves against risks deterrents such as price, quality, trust, and liquidity must be addressed (Platteau et al., 2017, Eling et al., 2014). Table 2.6 illustrates the extremely low uptake of microinsurance in the three regions, with health insurance faring slightly better than agricultural insurance. 

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = 'asis'}
library(tibble)
tribble(~ Region, ~ `Agricultural_Risks(%)`, ~ `Health_Risks(%)`,
        "Africa", "0.10", "0.74",
        "Latin America", "0.36", "1.24",
        "Asia and Oceania", "0.60", "0.74") %>% 
  
  kableExtra::kbl(., caption = "Uptake of Microinsurance", 
      
  booktabs = TRUE) %>% 
  
  kable_paper(full_width = FALSE, font_size = 8) %>% 
  
  footnote(general = "Platteau et al. (2017)",
           
  general_title = "Source: "
  ) 
  
```

Another product offered by MFIs is micro-pensions. Any pension scheme aims to provide some steady flow of income from retirement to death. Micro-pensions involve the medium and long-term savings schemes that generate capital which the scheme then invests in a variety of assets. The financial flows from these investments are then used to support the non-working elderly (Frankiewicz and Churchill, 2011). Pensions were devised in the western world during the industrial revolution when formal employment gradually replaced casual and self-employment. The challenge with micro-pensions is that most of the MFI clients are engaged in informal, mostly self-employment. It means, therefore that their cash flow is very uneven. Furthermore, most of the cash that the microentrepreneurs produce is prioritised towards the provision of the bare necessities of life, and not business growth or savings. Lastly, some MFIs are even involved in the distribution and marketing of clients output (Armendáriz and Morduch, 2010).

Although product diversification is beneficial to MFIs, it also poses several challenges. For example, some products may result in financial losses due to low demand or the lack of appropriate delivery channels, as in the case of micro-insurance (Platteau et al., 2017). Also, some products offered by MFIs may affect the uptake of another product, a form of cannibalisation (Elabed and Carter, 2015). Cannibalisation is frequent among the various classes of savings accounts, as the experience by Association for Social Advancement (ASA) in Bangladesh demonstrated (Campion and White, 1999, Meagher et al., 2006). Moreover, the failure of a product is likely to affect the image of the MFI among its clients. Furthermore, the diversity of products may affect the service quality as the staff’s attention is spread across more products. Lastly, the primary concern in MFIs is that the product diversification may cause mission drift (Christen and Cook, 2001, Chahine and Tannir, 2010). 

### Efficiency, Financial Performance, and Sustainability
\noindent The entry of shareholders and bondholders in addition to the traditional donors presumably increases the level of monitoring of the management (Stoughton and Zechner, 1998). From the agency theory perspective, monitoring is one of the means of resolving the agency conflicts. Also, bondholders and shareholders are likely to give MFIs an increased commercial orientation, through, for example, the expertise they bring in by way of the board of directors and the management team they hire (Lauer, 2008). 
Furthermore, some MFIs have adopted performance-based compensation schemes to address agency conflicts (Bateman, 2010a). Thus, any examination of the transformation of MFIs from lenses of the agency theory points towards improved efficiency and financial performance. Indeed, most studies have found a positive correlation between the injection of commercial capital and the financial performance of MFIs (Louis et al., 2013). For instance,  Daher and Le Saout (2013) find that transformed MFIs perform better financially. They note, however, that the better financial performance is inconsistent with social performance. 
Moreover, researchers have reservations regarding the improvements in financial performance by transformed MFIs. They point to the tendency of transformed MFIs to increase interest rates, engage in aggressive marketing and loan collection methods and offer incentives to managers based on volumes of loans (Ghosh, 2013). These tendencies may be detrimental to clients, lower portfolio quality, and leads to market saturation  (Hoque et al., 2011). 

On the other hand, there exists conflicting evidence on the effects of the transformation on social performance (and social efficiency), although increased commercial orientation may point to reduced social emphasis (Cobb et al., 2016, Daher and Le Saout, 2013). Other researchers find mixed outcomes (Abdulai and Tewari, 2017). It is then not clear whether the benefits of the transformation of MFIs accrue only to the providers of capital, clients, or to both parties. The inconsistent research output point to a significant research gap. Awaworyi Churchill and Nuhu (2016) pose an essential question, “What has failed? Microfinance or evaluation methods?”. The lack of consensus amongst researchers despite the abundance of research in the area indicates the need for more research.

### Enhanced Control and Governance 
\noindent The entry of commercially oriented shareholders and debt holders has the potential to enhance the control and governance of MFIs. First, commercial funders emphasise tightening the systems of internal controls and thus leading to improved management of the MFIs (Mersland and Strøm, 2010, Mersland and Strøm, 2009). Secondly, board representatives of commercial providers of capital are likely to uphold higher professional standards and possess the technical expertise to drive the organisation in the right direction. 
Also, the transformed MFI must conform with the local banking regulation, including the setting up of a board committee, management structures, disclosures, and stringent risk management procedures  (Frank et al., 2008). Likewise, the transformation also leads to the hiring of experienced human resource personnel that may have a bearing on the performance of the transformed MFI. Some of the research that uncovers a positive association between MFI transformation and improved financial performance attribute part of this improvement to enhanced control and governance. However, some research finds that the governance of MFIs affects neither their financial performance nor the outreach to the poor  (Mersland and Strøm, 2009).

2.10.6	Improved Customer Service
\noindent An essential consequence of the commercialisation of MFIs is customer orientation. One example of customer positioning is the development and provision of additional products like micro-insurance and microsavings (Chandrasekhar and Anantharaman, 2015). The new products are geared to meet the customer needs better. To increase customer convenience, most MFIs have adopted technology (Ouma et al., 2017). The fact that MFIs have done away with the one size fits all approach in product design is an illustration of increased customer focus. 

Importantly though is that the microfinance space has become more competitive with the entry of commercial banks that offer similar products (Kar, 2016). Also, the presence of shareholders and debtholders keen on a financial return means that the MFIs must strive towards retaining the existing customers and winning over new customers (Ballwieser et al., 2012). It means therefore that the customer experience is a critical differentiator in both the financial and social success of MFIs. Thus, the transformation of MFIs usually involves significant costs in the training of staff in customer care. 

\noindent Other Motives for the Transformation of MFIs
According to Christen and Cook (2001), there are several other reasons why MFIs transform into commercial entities. CGAP notes that legislative change may force an MFI to transform. For example, in Ghana, all MFIs are required to register with the Bank of Ghana (Bank of Ghana, 2016). Georgia enacted a law requiring NGOs to cease engaging in MF. Such a legislative requirement provides new institutional requirements for engaging in MF activities. MFIs also transform to gain legitimacy in the eyes of investors, customers, peer financial institutions and other lenders. The legitimacy would allow the transformed MFIs to access funds and raise additional capital with ease. Lastly, some MFIs have explicitly transformed to allow their employees or clients to become owners (Lauer, 2008). The move is designed to increase employee loyalty and minimise agency conflicts between employees and the management.

## Challenges of Transforming MFIs
\noindent The transformation of MFIs from NGOs to commercial entities poses several challenges. First, the transformation is costly- both financially and the in the time consumed (Frank et al., 2008). For instance, transformed MFIs incur higher costs arising from the increased reporting requirements and compliance with regulatory procedures (Meagher et al., 2006). The transformation also leads to increased spending on infrastructure (e.g., physical and logical security infrastructure) and human resources (e.g., lawyers, financial consultants, managers, and MFI personnel). Finally, the new products launched by transformed MFIs require extensive market research and marketing (Campion and White, 1999).

Also, legal and other administrative challenges plague the transformation process. In Kosovo, for example, there was a ruling that declared the transformation of MFIs into shareholder-owned funds unconstitutional (The Constitutional Court of the Republic of Kosovo, 2013). On the other hand, Georgia banned NGOs from offering MF services (Lauer, 2008). In the case of Georgia, the ban meant that the NGOs had to transform, spin-off their MF segments into separate business entities or abandon the MF business altogether. 
The administrative challenges arise out of three problems inherent in the conversion of MFIs into commercial stand-alone legal entities: integration into the formal financial system, ownership, and governance, and organisational development (Campion and White, 1999, Lauer, 2008, Mori and Mersland, 2014). The integration of the transformed MFIs into the financial system raises several challenges. First, in most jurisdictions, there were no specific regulations tailored for MFIs. Thus, MFIs transforming had little choice of an appropriate organisational structure (Campion and White, 1999). Most MFIs that transformed had to adopt the commercial banking structure with the attendant regulatory requirements, including the minimum capital. However, with the maturity of the regulatory framework, MFIs can choose from a variety of organisational forms, such as commercial MFIs, rural banks, commercial banks, or a variant of an NBFI. Several jurisdictions have created a legal framework specific to MFIs which has aided the transformation processes for MFIs. 

Before transformation, and even after, the move from the NGO setting to a commercial setting is a challenging process for most MFIs. For instance, the MFI must expend resources to develop commercial, organisational culture, acquire additional human resources and organise training programs to orient the existing staff members on the new business model. The development or acquisition and implementation of a management information system (MIS) suited for modern banking are especially pressing and costly procedures (Campion and White, 1999). For most MFIs that have moved from the NGO model, the culture of the organisation employees must be altered to cater for a more commercial, and thus, a more customer-centric culture. These issues relating to organisational development results in substantial costs of training and mentorship (Campion and White, 1999). 

Furthermore, the MFI transformation ushers in new equity and bondholders. The new ownership team involves international investors, employee stock ownership programs, and local private investors. The original NGO is usually part of the shareholding, although highly diluted (Périlleux et al., 2012). Most of the new owners require some form of financial return. In the case of K-Rep, identifying potential shareholders and negotiating a shareholder’s plan took a substantial amount of time (Mersland, 2009, Servin et al., 2012, Campion and White, 1999). 

Moreover, the new ownership structure requires that a board should be set up to oversee the running of the organisation. The board typically sets the mission and vision of the organisation and sets the investment strategy of the organisation. Setting up a useful board to steer an MFI through a transformation process is itself a challenge (Mori and Mersland, 2014). The new governance mechanism and the profit motive may create strains between the different stakeholders in the MFIs.

Also, the political-legal and social-economic environment pose a challenge to the transformation of MFIs, including the timing of the transformation. On the legal side, strict usury laws, tax legislation, and provisioning requirements may limit the financial sustainability of a transformed MFI (Meagher et al., 2006, Johnson, 2013). Like in the case of Georgia and Kosovo highlighted earlier, political-legal interventions may be the ultimate determinant for the decision to transform.  As the experience with K-Rep shows, the economic environment has a bearing on the valuation of assets and the speed of license processing by regulators.  For instance, the banking crisis in Kenya led to a delay in the licensing of K-Rep by about two years (Campion and White, 1999). The next chapter lays out the conceptual framework of the study. 



